\name: Deterministic and Coherent Implicits

### 2017-04-19 response.txt

* Presentation Improvements

  Thank you for the suggestions to improve the presentation.  We will
definitely take them into account. In particular we are very happy with the
comment on making a punchline at the end of Section 2 on stability as the key
idea; we agree that this will be very helpful to better convey the key idea to
readers.

  Tracking of variables, image what happens if we don't treat them spefically
  and consider this example resolution query

  a, a -> a |-_r forall b. b -> b

* Bugs in the typing rules

  We do indeed intentionally abuse the notation (we would not call
  that bugs), but are happy to reformulate as you suggest to clarify
  the intended meaning.

  Should we say something about Show?

* The intended type of bad on p. 5 is actually bad :: a -> a, 
  and not bad :: Trans a => a -> a, because we want to force
  the resolution of Trans at the definition of bad instead of
  the call sites of bad.

* Naming the implicit parameters only works if you want to directly
  refer to them. It does not work in the case of recursive resolution,
  e.g. Int => Bool, Int |- Bool, where multiple rules have to be combined.
  In such cases the system figures out how to combine the rules; the programmer
  does not have to write it himself.

* Provide example for \beta >_\tenv  \alpha

REVIEW 3

* We will give more practical examples of implicits, but this paper is mostly
  about restrictions to obtain coherence. The two cited papers

	B. C. d. S. Oliveira, A. Moors, and M. Odersky. 2010. Type classes as objects
	and implicits. In OOPSLA. ACM, New York, NY, USA.

	B. C. d. S. Oliveira, T. Schrijvers, W. Choi, W. Lee, and K. Yi. 2012.  e
	Implicit Calculus: A New Foundation for Generic Programming. In PLDI ’12. ACM,
	New York, NY, USA.

  provide many examples to motivate implicits.

### 2017-04-20 response.txt

* Presentation Improvements

  Thank you for the suggestions to improve the presentation.  We will
  definitely take them into account. In particular we are very happy with the
  comment on making a punchline at the end of Section 2 on stability as the key
  idea; we agree that this will be very helpful to better convey the key idea to
  readers.

  We would like to clarify the need for tracking the type variables referred
  to on p.13. Consider for example resolving the type

		ρ = b.b → b

  against the environment

  		Γ0 = ∀c.c → c ~> id, a, a → a ~> f

  i.e.

	     	Γ0 ⊢ ρ

  Imagine what would happen if we did not track the type variables. Firstly,
  using rule R-TAbs we would extend the environment with b

  		Γ1 = Γ0, b

  and resolve the simple type

	     	Γ1 ⊢ b → b

  which would get stuck on a → a when considering which rule matches in Γ1. In particular,
  L-RuleMatch does not apply because a → a is different from b → b. Also L-RuleNoMatch does
  not apply because a → a and b → b unify under the substitution [b := a]; thus skipping 
  rule a → a would be deemed unstable.

  However, there is an error in the above reasoning. Code inlining cannot
  result in b being instantiated to a, because b is not free in the environment 
  Γ0, which is the environment at the point in the code where the query happens.
  To avoid this reasoning error, we keep track of the variables that are free
  in the environment in Γ0. These are the variables that can be substituted, while
  the variables added to the environment by rule R-TAbs cannot be substituted.

  We will integrate this example in the text.

### 2017-04-20 response.txt

  TODO: Should we say something about Show?

* The intended type of bad on p. 5 is actually bad :: a -> a, 
  and not bad :: Trans a => a -> a, because we want to force
  the resolution of Trans at the definition of bad instead of
  the call sites of bad.

* Naming the implicit parameters only works if you want to directly
  refer to them. It does not work in the case of recursive resolution,
  e.g. Int => Bool, Int |- Bool, where multiple rules have to be combined.
  In such cases the system figures out how to combine the rules; the programmer
  does not have to write it himself.

* The β >Γ α judgement is related to the tracking of variables and the checking of
  stability discussed above. 
  It performs another necessary sanity check: a substitution [α := β] (*) due to
  inlining is only possible if β is in scope of α. This property is verified in
  terms of the relative positions of α and β in the environment: the latter must
  appear before the former.

  (*) This generalizes to substitutions [α := σ] where β is a free variable in σ.

### 2017-04-20 response.txt

  However, there is an error in the above reasoning. Code inlining cannot
  result in b being instantiated to a, because b is not free in the environment 
  Γ0, which is the environment at the point in the code where the query happens.
  To avoid this reasoning error, we keep track of the variables that are free
  in the environment in Γ0. These are the variables that can be substituted, while
  the variables added to the environment by rule R-TAbs cannot be substituted.

  We will integrate this example in the text.

* Bugs in the typing rules

  We do indeed intentionally abuse the notation (although we would not call
  that bugs), but are happy to reformulate as you suggest to clarify
  the intended meaning.

@@

  TODO: Should we say something about Show?

* The intended type of bad on p. 5 is actually bad :: a -> a, 
  and not bad :: Trans a => a -> a, because we want to force
  the resolution of Trans at the definition of bad instead of
  the call sites of bad.

* Naming the implicit parameters only works if you want to directly
  refer to them. It does not work in the case of recursive
  resolution. For example consider the program:

  \(Int => Bool) . \Int . ?Bool

  where multiple rules have to be combined to resolve ?Bool.
  In such cases the system figures out how to combine the rules; the programmer
  does not have to write the composition himself.

@@

* TODO: Provide example for \beta >_\tenv  \alpha

REVIEW 3

* Our work is on an improvement to the implicit calculus. Our
  improvement is mainly about working out the restrictions needed to
  ensure coherence. So it does not add any new programming features
  per se. Practical examples illustrating the uses of
  implicits were already presented in the following two cited papers:

@@

	B. C. d. S. Oliveira, A. Moors, and M. Odersky. 2010. Type classes as objects
	and implicits. In OOPSLA. ACM, New York, NY, USA.

	B. C. d. S. Oliveira, T. Schrijvers, W. Choi, W. Lee, and K. Yi. 2012.  The
	Implicit Calculus: A New Foundation for Generic Programming. In PLDI ’12. ACM,
	New York, NY, USA.

  Thus our paper focuses only on coherence and examples about
  coherence.

### 2017-04-20 response.txt

* Naming the implicit parameters only works if you want to directly
  refer to them. It does not work in the case of recursive
  resolution. For example consider the program:

  \(Int => Bool) . \Int . ?Bool

  where multiple rules have to be combined to resolve ?Bool.
  In such cases the system figures out how to combine the rules; the programmer
  does not have to write the composition himself.

* The β >Γ α judgement is performs another necessary sanity check during unification.
  Consider for example resolving the type

		ρ = (∀β.β → β) → Int

  against the environment with a single rule

  		Γ0 = ∀α.(∀β.β → α) → Int

  This will lead in rule Alg-M-Simp to the unification

		mgu_{Γ1,α}(∀α.(∀β.β → α) → Int,(∀β.β → β) → Int)

  with

  		Γ1 = Γ0, α

  After several steps, we get

		mgu_{Γ2,α}(α,β)

  with

  		Γ2 = Γ0, α, β

  The result here cannot be [β := α] because β is universally quantified
  and thus cannot be instantiated. This is why rule U-InstR should not apply;
  the mechanism used is that mgu is indexed in the unifiable variables --   
  only α in this example.

  However, we also cannot have [α := β] because β is not in scope at the point
  where α has to be instantiated. We see this because β occurs after α in the 
  environment Γ2. This scope check is captured in judgement β >Γ2 α. This check
  is the one that prevents rule U-InstL from applying and from producing an invalid
  substitution.

### 2017-04-20 response.txt

	     	Γ1 ⊢ b → b

  which would get stuck on a → a when considering which rule matches in Γ1. In particular,
  L-RuleMatch does not apply because a → a is different from b → b. Also L-RuleNoMatch does
  not apply because a → a and b → b unify under the substitution [b := a]; thus skipping 
  rule a → a would be deemed unstable.

  However, there is an error in the above reasoning. Code inlining cannot
  result in b being instantiated to a, because b is not free in the environment 
  Γ0, which is the environment at the point in the code where the query happens.
  To avoid this reasoning error, we keep track of the variables that are free
  in the environment in Γ0. These are the variables that can be substituted, while
  the variables added to the environment by rule R-TAbs cannot be substituted.

  In the end, by tracking the free variables, we can accept the
  resolution above, whereas without such tracking the program would be rejected.

@@

* Naming the implicit parameters only works if you want to directly
  refer to them. It does not work in the case of recursive
  resolution. For example consider the program:

  \(Int => Bool) . \Int . ?Bool

  where multiple rules have to be combined to resolve ?Bool.
  In such cases the system figures out how to combine the rules; the programmer
  does not have to write the composition himself.

* The β >Γ α judgement performs a necessary sanity check during
  unification, which is useful to prevent some invalid unifications
  that would arise due to scoping issues.
  Consider for example resolving the type

@@

	B. C. d. S. Oliveira, A. Moors, and M. Odersky. 2010. Type classes as objects
	and implicits. In OOPSLA. ACM, New York, NY, USA.

	B. C. d. S. Oliveira, T. Schrijvers, W. Choi, W. Lee, and K. Yi. 2012.  The
	Implicit Calculus: A New Foundation for Generic Programming. In PLDI ’12. ACM,
	New York, NY, USA.

  Thus our paper focuses only on coherence and examples about
  coherence.

  Nevertheless we will try our best to make the presentation more
  acessible by taking all the suggestions into account.

### 2017-04-20 response.txt

  and resolve the simple type

	     	Γ1 ⊢ b → b

  which would get stuck on a → a when considering which rule matches in Γ1. In particular,
  L-RuleMatch does not apply because a → a is different from b → b. Also L-RuleNoMatch does
  not apply because a → a and b → b unify under the substitution [b := a]; thus skipping 
  rule a → a would be deemed unstable.

  However, there is an error in the above reasoning. Code inlining cannot
  result in b being instantiated to a, because b is not in scope
  at the point in the code where the query happens (i.e., b does not appear in Γ0).  
  To avoid this reasoning error, we keep track of the variables that appear
  in the environment Γ0. These are the variables that can be substituted, while
  the variables added to the environment by rule R-TAbs cannot be substituted.

@@

	B. C. d. S. Oliveira, A. Moors, and M. Odersky. 2010. Type classes as objects
	and implicits. In OOPSLA. ACM, New York, NY, USA.

	B. C. d. S. Oliveira, T. Schrijvers, W. Choi, W. Lee, and K. Yi. 2012.  The
	Implicit Calculus: A New Foundation for Generic Programming. In PLDI ’12. ACM,
	New York, NY, USA.

  Thus our paper focuses only on coherence and examples about
  coherence.

  Nevertheless we will do our best to make the presentation more
  accessible by taking all the suggestions into account.

### 2017-06-06 Main.tex
comments on Section 3
authoryear citations
add conclusion
jfp version: first port

\documentclass[acmlarge, anonymous, review]{acmart}
%\documentclass[prodmode,acmtoplas]{acmsmall}

@@

\newcommand{\gbox}[1]{\colorbox{light}{$\!\!#1\!\!$}}

@@

% Document starts
\begin{document}

% Page heads
\markboth{B. Oliveira et al.}{The Implicit Calculus: A New Foundation for Generic Programming}

% Title portion
\title{\name: Deterministic and Coherent Implicits}
\author{Ommitted for anonymous submission}

@@

\begin{comment}
\category{D.3.2}{Programming Languages}   
                {Language Classifications}
                [Functional Languages, Object-Oriented Languages]
\category{F.3.3}{Logics and Meanings of Programs}   
                {Studies of Program Constructs}
                []

\terms{Languages}

\keywords{
Implicit parameters, type classes, C++ concepts, generic programming,
Haskell, Scala}

\acmformat{Oliveira, B. C. d. S., Schrijvers, T., Choi, W., Lee, W., Yi, K., Wadler, P.
201?. The implicit calculus: a new foundation for generic programming.}

\begin{bottomstuff}
Author's addresses: 
B. C. d. S. Oliveira, Department of Computer Science, Hong Kong University; 
T. Schrijvers, Department of Applied Mathematics and Computer Science, Ghent University;
W. Choi {and} W. Lee {and} K. Yi, \ldots, Seoul National University;
P. Wadler, \ldots, University of Edinburgh.
\end{bottomstuff}
\end{comment}

\maketitle

@@

% Bibliography
\bibliographystyle{ACM-Reference-Format}
\bibliography{papers}

% Appendix
\appendix
\section*{APPENDIX}
\setcounter{section}{1}

>>>>

\documentclass[acmlarge, authorname]{acmart}
%\documentclass[prodmode,acmtoplas]{acmsmall}

@@

\newcommand{\gbox}[1]{\colorbox{light}{$\!\!#1\!\!$}}
\newcommand{\plw}[1]{[{\color{blue}\text{PLW:}} #1]}

@@

% Document starts
\bibliographystyle{ACM-Reference-Format}
\citestyle{acmauthoryear}
\begin{document}

% Page heads
% Title portion
\title{\name: Deterministic and Coherent Implicits}
\author{Tom Schrijvers}
\affiliation{KU Leuven}
\author{Bruno C.d.S. Oliveira}
\affiliation{The University of Hong Kong}
\author{Philip Wadler}
\affiliation{University of Edinburgh}

@@

\setcopyright{none}
\maketitle

@@

\newpage
% Bibliography
\bibliography{papers}

\newpage
% Appendix
\appendix
\section*{APPENDIX}
\setcounter{section}{1}
